{\helveticaitalic{Supplementary Material}}: \\ \helvetica{Xolotl: An Intuitive and Approachable Neuron \& Network Simulator in MATLAB}

### 2018-08-08 paper/paper.tex

Single-compartment Hodgkin-Huxley-like models with three conductances and stomatogastric ganglion neuron models with eight conductances were generated using conductance dynamics from \cite{liuModelNeuronActivityDependent1998}) in the simulation environments. "xolotl" uses the exponential Euler method for integrating membrane potential (\cite{dayanTheoreticalNeuroscience2001}). "DynaSim" was implemented with a 2\textsuperscript{nd}-order Runge-Kutta integration scheme as recommended for high-performance in the documentation. "NEURON" used an implicit Euler regime (\cite{hinesNEURONSimulationEnvironment1997}).

@@

DESCRIBE BENCHMARK RESULTS

@@

\begin{figure}
	\centering
	\includegraphics[width=1.0\linewidth]{gfx/figure_HH}
	\caption{Setting up a single-compartment model and injecting current into it. (A) Schematic representation of a single-compartment neuron with three populations of ion channels (colored rectangles). (B) In the simulator, the soma is represented using an object of class $compartment$ and the populations of ion channels are represented by $conductance$ objects contained within the compartment object. (C) The code snippet shown sets up this neuron model, injects current, integrates and plots the voltage, and displays activation functions, all in a few lines of code. (D) Simulated voltage trace of a Hodgkin-Huxley model with three conductances and 0.2 nA of injected current. Colors indicate the dominant current (gold is fast sodium (NaV), blue is delayed rectifier (Kd), red is Leak). (E) firing rate $vs$. current (f-I) curve of this neuron. (F-G) Steady-state gating functions for activation (m) and inactivation (h) gating variables. (H-I) Voltage-dependence of time constants for activation (m) and inactivation (h) gating variables}
	\label{fig:figurehh}
\end{figure}

\begin{figure}[!htb]
	\centering
	\includegraphics[width=1.0\linewidth]{gfx/figure_clamp}
	\caption{Simulating a voltage-clamp experiment. (A) The diagram shows a cell with delayed rectifier potassium conductance (from \cite{liuModelNeuronActivityDependent1998}) that is being recorded from in two-electrode voltage clamp. (B) The code snippet shown here sets up a model with a single compartment and a single channel type, and clamps the cell to a constant voltage and integrates it. (C) Voltage steps that the cell is clamped to. (D) Current injected into the cell required to clamp it to the voltages in (C). (E) Asymptotic clamped current $vs.$ clamped voltage for this cell. (F) Accounting for the reversal potential of Potassium ions yields the conductance-voltage curve of this channel type. (G) Normalized conductance-voltage curves, with sigmoid fits with various exponents. (H) An exponent of $n=4$ yields the best fit, allowing for the characterization of the activation function of this channel type. }
	\label{fig:figureclamp}
\end{figure}

@@

\begin{figure}[!htb]
	\centering
	\includegraphics[width=1.0\linewidth]{gfx/figure_snapshot}
	\caption{Snapshots allow the user to bookmark points in parameter and state space of the model and return to them $ad$   $arbitrium$. The initial state (orange disc) of the single compartment model in the previous example is saved using the \texttt{snapshot} method. This method saves all parameters and dynamic variables of the model in a named state. The first column shows the profile of conductances and the voltage dynamics of the model at this point in time. The maximal conductances of the Calcium channels are set to zero (purple), changing the voltage dynamics of the neuron (purple plots). After evolving the model for some time (green), the conductance profile and voltage dynamics returns to a state similar to the initial state (green plots). This configuration is now saved in a state called $final$ and the initial configuration is returned to using the \texttt{reset} method (backwards arrow from green to orange). Another parameter is changed (the Calcium target), and the model is integrated to reach a new state (blue disc) where the voltage dynamics are different from the $initial$ state. In summary, any state can be bookmarked using a descriptive name using the \texttt{snapshot} method, and can be returned to using the \texttt{reset} method.}
	\label{fig:figuresnapshot}
\end{figure}

@@

\begin{figure}
	\centering
	\includegraphics[width=1.0\linewidth]{gfx/figure_manipulate}
	\caption{Manipulating neuron parameters in real time. Any set of parameter in the model can be manipulated; here, the maximal conductance of every conductance type in the model from Figure 1 is being manipulated using the code snippet shown here. The screenshot shows a GUI with sliders for every parameter of interest that is created by the \texttt{manipulate} method, and linked to an arbitrary number of visualization functions. In this example, two visualization functions are used: the built in \texttt{plot} method (A) and a custom function that computes the firing-rate-$vs$.-injected current curve for this neuron (B). Both plots refresh themselves with every movement of any slider, allowing the user to build intuition about how every parameter controls the dynamical behavior of the model. A screen recording of this model being manipulated in real time is included in Supplementary Material XX. }
	\label{fig:figuremanipulate}
\end{figure}

@@

\begin{figure}
	\centering
	\includegraphics[width=1.0\linewidth]{gfx/figure_benchmark}
	\caption{\texttt{xolotl} benchmarked against \texttt{DynaSim} and \texttt{NEURON}. The top row benchmarks using a tonically-firing Hodgkin-Huxley model with 3 conductances with constant injected current. The bottom row repeats the same benchmarks using a bursting stomatogastric ganglion model with 8 conductances. (A,E) Ratio of run-time to simulation time (speed factor) over increasing integration time step. (B,F) Error over increasing integration time step as computed by the LeMasson matrix where the baseline condition is the voltage trace at $\Delta t = 0.001$ ms. (C,G) Simulations over increasing simulation time where $\Delta t = 0.1$ ms. (D,H) Simulations over increasing number of unconnected compartments.}
	\label{fig:figurebenchmark}
\end{figure}

>>>>

Single-compartment Hodgkin-Huxley-like models with three conductances and stomatogastric ganglion neuron models with eight conductances were generated using conductance dynamics from \cite{liuModelNeuronActivityDependent1998}) in the simulation environments. "xolotl" uses the exponential Euler method for integrating membrane potential (\cite{dayanTheoreticalNeuroscience2001}). "DynaSim" with "C"-compilation was implemented with a 2\textsuperscript{nd}-order Runge-Kutta integration scheme as recommended for high-performance in the documentation. "NEURON" used an implicit Euler regime (\cite{hinesNEURONSimulationEnvironment1997}).

@@

"xolotl" performed well in the benchmark simulations. Over increasing time step, "xolotl" was faster than "DynaSim" and "NEURON" with smaller error. During longer simulations "xolotl" and "DynaSim" increased in speed. This is likely due to the decreased impact of set-up overhead from "MATLAB", an interpreted language. For most simulation times, "xolotl" is roughly ten times faster than "DynaSim" and about twice as fast in the long-simulation time limit. "NEURON" performed consistently at high speed, likely due to a lack of programmatic overhead, though "DynaSim" and "xolotl" were significantly faster at large time steps and during long simulations.

@@

\begin{figure}
	\centering
	\includegraphics[width=1.0\linewidth]{gfx/figure_HH}
	\caption{Setting up a single-compartment model and injecting current into it. (A) Schematic representation of a single-compartment neuron with three populations of ion channels (colored rectangles). (B) In the simulator, the soma is represented using an object of class \texttt{compartment} and the populations of ion channels are represented by \texttt{conductance} objects contained within the compartment object. (C) The code snippet shown sets up this neuron model, injects current, integrates and plots the voltage, and displays activation functions, all in a few lines of code. (D) Simulated voltage trace of a Hodgkin-Huxley model with three conductances and 0.2 nA of injected current. Colors indicate the dominant current (gold is fast sodium (NaV), blue is delayed rectifier (Kd), red is Leak). (E) firing rate \vs. current (f-I) curve of this neuron. (F-G) Steady-state gating functions for activation (m) and inactivation (h) gating variables. (H-I) Voltage-dependence of time constants for activation (m) and inactivation (h) gating variables}
	\label{fig:figurehh}
\end{figure}

\begin{figure}[!htb]
	\centering
	\includegraphics[width=1.0\linewidth]{gfx/figure_clamp}
	\caption{Simulating a voltage-clamp experiment. (A) The diagram shows a cell with delayed rectifier potassium conductance (from \cite{liuModelNeuronActivityDependent1998}) that is being recorded from in two-electrode voltage clamp. (B) The code snippet shown here sets up a model with a single compartment and a single channel type, and clamps the cell to a constant voltage and integrates it. (C) Voltage steps that the cell is clamped to. (D) Current injected into the cell required to clamp it to the voltages in (C). (E) Asymptotic clamped current \vs clamped voltage for this cell. (F) Accounting for the reversal potential of Potassium ions yields the conductance-voltage curve of this channel type. (G) Normalized conductance-voltage curves, with sigmoid fits with various exponents. (H) An exponent of $n=4$ yields the best fit, allowing for the characterization of the activation function of this channel type. }
	\label{fig:figureclamp}
\end{figure}

@@

\begin{figure}[!htb]
	\centering
	\includegraphics[width=1.0\linewidth]{gfx/figure_snapshot}
	\caption{Snapshots allow the user to bookmark points in parameter and state space of the model and return to them \textit{ad arbitrium}. The initial state (orange disc) of the single compartment model in the previous example is saved using the \texttt{snapshot} method. This method saves all parameters and dynamic variables of the model in a named state. The first column shows the profile of conductances and the voltage dynamics of the model at this point in time. The maximal conductances of the Calcium channels are set to zero (purple), changing the voltage dynamics of the neuron (purple plots). After evolving the model for some time (green), the conductance profile and voltage dynamics returns to a state similar to the initial state (green plots). This configuration is now saved in a state called \textit{final} and the initial configuration is returned to using the \texttt{reset} method (backwards arrow from green to orange). Another parameter is changed (the Calcium target), and the model is integrated to reach a new state (blue disc) where the voltage dynamics are different from the \textit{initial} state. In summary, any state can be bookmarked using a descriptive name using the \texttt{snapshot} method, and can be returned to using the \texttt{reset} method.}
	\label{fig:figuresnapshot}
\end{figure}

@@

\begin{figure}
	\centering
	\includegraphics[width=1.0\linewidth]{gfx/figure_manipulate}
	\caption{Manipulating neuron parameters in real time. Any set of parameter in the model can be manipulated; here, the maximal conductance of every conductance type in the model from Figure 1 is being manipulated using the code snippet shown here. The screenshot shows a GUI with sliders for every parameter of interest that is created by the \texttt{manipulate} method, and linked to an arbitrary number of visualization functions. In this example, two visualization functions are used: the built in \texttt{plot} method (A) and a custom function that computes the firing-rate-\vs-injected current curve for this neuron (B). Both plots refresh themselves with every movement of any slider, allowing the user to build intuition about how every parameter controls the dynamical behavior of the model. A screen recording of this model being manipulated in real time is included in Supplementary Material. }
	\label{fig:figuremanipulate}
\end{figure}

@@

\begin{figure}
	\centering
	\includegraphics[width=1.0\linewidth]{gfx/figure_benchmark}
	\caption{\texttt{xolotl} benchmarked against \texttt{DynaSim} and \texttt{NEURON}. The top row benchmarks using a tonically-firing Hodgkin-Huxley model with 3 conductances with constant injected current. The bottom row repeats the same benchmarks using a bursting stomatogastric ganglion model with 8 conductances. (A,E) Ratio of run-time to simulation time (speed factor) over increasing integration time step. (B,F) Error over increasing integration time step as computed by the LeMasson matrix where the baseline condition is the voltage trace at $\Delta t = 0.001$ ms. (C,G) Simulations over increasing simulation time where $\Delta t = 0.1$ ms. (D,H) Simulations over increasing number of unconnected compartments in the same network.}
	\label{fig:figurebenchmark}
\end{figure}

### 2018-08-08 paper/supplementary/supplementary.tex

PUT FIGURE HERE

>>>>

Bursting models from the Prinz stomatogastric ganglion model database (\cite{prinzAlternativeHandtuningConductancebased2003}) were simulated for 15 seconds at varying time resolution in order to characterize the stability of the exponential Euler numerical integration method. The exponential Euler method is generally accurate up to $\approx 0.1$ ms for most models tested. A time step of $\approx 0.05$ ms gives a good compromise between accuracy and speed (\cite{prinzAlternativeHandtuningConductancebased2003, prinzSimilarNetworkActivity2004}).

\begin{figure}
\centering
\includegraphics[width=1.0\linewidth]{gfx/figure_sup_stability}
\caption{Stability of exponential Euler method using models randomly chosen from the Prinz database. (A, C, E) Three metrics of bursting neurons: burst frequency, average spikes per burst, and duty cycle were computed over increasing time step. (B, D, F) The metrics were normalized to the first value in for each model. A line of zero slope would indicate no significant change in integration error over increasing time step. Each line indicates a model with different maximal conductances.}
\label{fig:figuresupstability}
\end{figure}

\cleardoublepage
